\section{Initial conditions for Moose objects}
\label{chap:IC-moose-objects}

For definition of the initial (starting) conditions for the variables a Moose
simulation the \href{https://mooseframework.inl.gov/syntax/ICs}{ICs system} is
to be used.

(to be written)

\section{Gravitation and initial stress state}
\label{chap:IC-stress-state}

The initial stress state of geotechnical simulations is often non-trivial. This
is because the section of geosphere to be modelled often has a history of
millions of years under different loading conditions, including gravity and
tectonic processes. In addition, the Earth's surface is often non-horizontal,
so that stress trajectories near the surface are oriented accordingly.
Furthermore, the initial stress state may be anisotropic with a specific
orientation. Gravity and the initial stress state are normally taken into
account by a combination of the following elements:

\begin{itemize}
  \item Definition of appropriate displacement boundary conditions for the model (see
        \autoref{chap:model-configuration-boundary-fixities}).
  \item Activation of gravitational body force (see
        \autoref{chap:IC-stress-state-gravity}).
  \item Definition of eigenstrains from the initial stress field (e.g. using
        \href{https://mooseframework.inl.gov/source/materials/ComputeEigenstrainFromInitialStress.html}{ComputeEigenstrainFromInitialStress},
        see \autoref{chap:IC-stress-state-simple} and
        \autoref{chap:IC-stress-state-anisotropic}).
\end{itemize}

\importantbox{
  In simulations including pore water, ‘effectice stresses’ are to be used.
  This includes the definition of the initial stress field, where also
  effective stresses are to be given.
}

\subsection{Gravity}
\label{chap:IC-stress-state-gravity}

To consider gravity, the following aspects must be included in the model:
\begin{itemize}
  \item When compiling the MooseApp: As the effect of gravity is taken into account in
        interaction with the material density, the Moose module \codeword{MISC} must be
        active in addition to the Moose module \codeword{SOLID_MECHANICS}.
  \item Gravity must be activated in the input file. This is done by adding a special
        kernel of the type \codeword{Gravity} as shown in
        \autoref{initial-conditions-gravity}. The direction and strength of the
        gravitational field is to be defined here.
  \item The materials must be assigned a material density
        (\autoref{initial-conditions-density}).
\end{itemize}

\begin{lstlisting}[language=perl, float, caption={Gravity kernel in a Moose inut file},label={initial-conditions-gravity}]
[Kernels]
    [gravity]
        type = Gravity
        use_displaced_mesh = false
        variable = disp_z           # the displacement variable the gravity is associated with
        value = -9.81               # in m/s^2
    []
[]
\end{lstlisting}

\begin{lstlisting}[language=perl, float, caption={Assignment of a density to subdomain ‘block1’},label={initial-conditions-density}]
[Materials]
    [undrained_density]
      type = GenericConstantMaterial
      block = 'Block1'
      prop_names = density
      prop_values = 2500            # in kg/m^3
    []
[]
\end{lstlisting}

{\hfuzz=20pt
\subsection{Initial stress state using ComputeEigenstrainFromInitialStress}
}
\label{chap:IC-stress-state-simple}

In a uninfluenced homogeneous horizontal half-space (or a section thereof)
under the effect of gravity, the initial stress state corresponds to a
‘geostatic stress field’. This geostatic stress field can be represented
analytically for a stress point as follows:

\begin{equation}
  \sigma'_{ini,vert}=\sigma'_{ini,vert,h=0}+\gamma' h
\end{equation}
\begin{equation}
  \sigma'_{ini,horiz}=\sigma'_{ini,vert} K_0 \quad \text{with } \quad K_0 = 1-\sin\varphi
\end{equation}

In these equations, $\gamma'=\rho' g$ corresponds to the weight of the soil,
$h$ to the overburden height, $\sigma'_{ini,vert,h=0}$ to the effective
vertical stress at $h=0$ and $K_0$ to the earth pressure coefficient. This
earth pressure coefficient is usually estimated as a function of the angle of
internal friction $\varphi$ (simplified equation according to Jaky).

\autoref{initial-conditions-ComputeEigenstrainFromInitialStress} shows the
definition of a geostatic initial stress state using two functions
\codeword{ini_xx_yy} and \codeword{ini_zz} and their subsequent use in a
\codeword{ComputeEigenstrainFromInitialStress}.

\begin{lstlisting}[language=perl, float, caption={Definition of a geostatic initial stress state using ‘ComputeEigenstrainFromInitialStress’ },label={initial-conditions-ComputeEigenstrainFromInitialStress}]
[Functions]
    [ini_xx_yy]   # function describing the initial effective horizontal stress field
      type = ParsedFunction
      vars = 'sig_top   z_top   rho      g    k0 '
      vals = '-1.5      0       0.0025   10   0.3'
      value = '(sig_top - rho * g * (z_top - z)) * k0'
    []
    [ini_zz]      # function describing the initial effective vertical stress field
      type = ParsedFunction
      vars = 'sig_top   z_top   rho      g'
      vals = '-1.5      0       0.0025   10'
      value = '(sig_top - rho * g * (z_top - z))'
    []
[]
  
[Materials]
    [strain]
      type = ComputeIncrementalSmallStrain
      volumetric_locking_correction = false
      eigenstrain_names = ini_stress        # pointing to the initial stress field
    []
    [ini_stress]   # initial stress field (effective stresses)
      type = ComputeEigenstrainFromInitialStress
      eigenstrain_name = ini_stress
      initial_stress = 'ini_xx_yy 0 0   0 ini_xx_yy 0   0 0 ini_zz'
    []
[]
\end{lstlisting}

\subsection{Initial stress state for anisotropic stresses}
\label{chap:IC-stress-state-anisotropic}

Align the model to the orientation of the initial stresses

(to be written)

\subsection{Checking the initial stress state}
\label{chap:IC-check}

In geotechnics, the initial state including the initial stress state should be
in equilibrium. Therefore, in a first time step without any changes to the
model ‘nothing’ should happen (only neglectable deformations and changes in the
stress field).

This can be checked by introducing such a time step without any changes as a
first time step and assessing the system response (e.g. in Paraview).
